Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-12 paper_io_sim.tex
Update commented text

            As our focus is on I/O rather than compute, we measured
            application task CPU times on a cluster node
            (Table~\ref{table:cputime}), and used these durations in our
            simulations. For the Python simulator, we used CPU times
            directly in the simulation. For \wrench and \wrench-cache, we
            determined the corresponding number of flops on a 1~Gflops CPU
            and used these values in the simulation. The final simulated
            platform and application used in our experiments are avaiable at
            \url{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}.

@@

            Our fourth and final experiment (\textit{Exp Nighres})
            simulates the Nighres CRUISE cortical reconstruction workflow
            \tristan{describe data, and if the execution was single node or
            distributed.}

@@

        Simulated memory profiles are highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). With 20~GB files, memory
        profiles almost exactly match the real ones, although dirty data
        seems to be flushing faster in real life than in simulation
        \tristan{any idea why?}, a behavior also observed with 100~GB files. With
        100~GB files, used memory reaches total memory during the first write,
        triggering dirty data flushing. Used memory drops to cached memory
        when application tasks release anonymous memory. Simulated cached
        memory is highly consistent with real values, except toward the end
        of Read 3 where it slightly increases in simulation but not in
        reality. Our explanation could be that we allocate anonymous memory 
        chunk-by-chunk to application, while anonymous memory is allocated 
        when as soon as task starts.
        As expected, dirty data remains under the
        dirty ratio. The Python prototype and \wrench-cache exhibit nearly
        identical memory profiles, which reinforces the confidence in our
        implementation.

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, since all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, 
        only a part of it is cached. 
        This is due to the fact that the in simulation, data is flushed and evicted 
        from the inactive list first, then from the active list. Thus, File 2, which is 
        accessed again by Read 2 and is cached in the active list, is not flushed and evicted, 
        while File 3 is flushed and evicted since it is accessed once only and cached 
        in the inactive list. This may not correctly simulate what really happens 
        in reality.

@@

            Page cache simulation importantly reduces the simulation error
            on remote storage as well (Figure~\ref{fig:multi_nfs}). This
            manifests only for reads, as only writethrough cache was
            enabled in the NFS server. Both \wrench and \wrench-cache
            underestimate write times due to the discrepancy between
            simulated and real bandwidths mentioned previously. For reads,
            this discrepancy only impacts the results beyond 22
            applications since before this threshold, most reads result in cache
            hits \tristan{Dzung, pls check that.}.

@@

    \section{Conclusion}
    \label{discussion}
        Page cache undoubtedly has important impacts on performance, in
        particular for data-intensive applications. Numerical simulation is
        an effective approach to predict and compare the performance of
        applications, schedulers and infrastructures. However, page cache
        is missing in many simulation tools for distributed systems. In
        this study, we proposed a simulation model of I/Os with page cache,
        we implemented it in the \simgrid-based \wrench library, and we
        evaluated its benefits in different conditions. Results show that
        our model very substantially improves simulation accuracy,
        decreasing absolute relative errors between simulated and real
        performance by a factor close to 10 (see results of the
        single-threaded experiment). Furthermore, this level of accuracy
        can be achieved without a fine-grained level of details
        \tristan{not sure what you mean}. Our page cache model is available
        in \wrench from commit \tristan{abcd}, ready to be used for the
        simulation of Big Data applications.

        Simulation results could be made even more accurate by a deeper
        investigation of cache eviction mechanisms \tristan{be more
        specific}. The availability of asymmetrical bandwidths in the
        forthcoming \simgrid release will also directly improve our
        simulation accuracy. Relevant future work includes simulating
        application memory consumption, random I/O, file readahead and
        persistent disk storage.

>>>>

            As our focus is on I/O rather than compute, we measured
            application task CPU times on a cluster node
            (Table~\ref{table:cputime}), and used these durations in our
            simulations. For the Python simulator, we used CPU times
            directly in the simulation. For \wrench and \wrench-cache, we
            determined the corresponding number of flops on a 1~Gflops CPU
            and used these values in the simulation. The final simulated
            platform and application used in our experiments are available at
            commit \href{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}{ec6b43561b}.

@@

            Our fourth and final experiment (\textit{Exp Nighres})
            simulates the Nighres CRUISE cortical reconstruction workflow, which consists 
            of four steps. The first two steps are skull stripping and MGDM segmentation, 
            which are in this Nighres \href{{https://nighres.readthedocs.io/en/latest/auto_examples/example_01_tissue_classification.html}}{tissue classification example}. 
            The last two tasks are region extraction and CRUISE cortical reconstruction 
            in the   
            \href{https://nighres.readthedocs.io/en/latest/auto_examples/example_02_cortical_depth_estimation.html}{cortical depth estimation example}.
            \tristan{describe data, and if the execution was single node or
            distributed.}

@@

        Simulated memory profiles are highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). With 20~GB files, memory
        profiles almost exactly match the real ones, although dirty data
        seems to be flushing faster in real life than in simulation
        \tristan{any idea why?}, a behavior also observed with 100~GB files. With
        100~GB files, used memory reaches total memory during the first write,
        triggering dirty data flushing. Used memory drops to cached memory
        when application tasks release anonymous memory. Simulated cached
        memory is highly consistent with real values, except toward the end
        of Read 3 where it slightly increases in simulation but not in
        reality. Our explanation could be that in reality, because File 3 is
        entirely in cache, the cache amount does not increase when File 3 is read.
        While in simulation, a part of File 3 is read from disk, leading to an increase
        in cache amount followed by a decrease when eviction is called.
        Apart from that, dirty data remains under the dirty ratio as expected. 
        The Python prototype and \wrench-cache exhibit nearly
        identical memory profiles, which reinforces the confidence in our
        implementation.

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, since all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, 
        only a part of it is cached. 
        This is due to the fact that the in simulation, data is flushed and evicted 
        from the inactive list first, then from the active list. Thus, File 2, which is 
        accessed again by Read 2 and is cached in the active list, is not flushed and evicted, 
        while File 3 is flushed and evicted since it is accessed once only and cached 
        in the inactive list. 
        This flushing and eviction order in simulation may not correctly reflect the 
        order in reality and lead to the difference.

@@

            Page cache simulation importantly reduces the simulation error
            on remote storage as well (Figure~\ref{fig:multi_nfs}). This
            manifests only for reads, as only writethrough cache was
            enabled in the NFS server. Both \wrench and \wrench-cache
            underestimate write times due to the discrepancy between
            simulated and real bandwidths mentioned previously. For reads,
            this discrepancy only impacts the results beyond 22
            applications since before this threshold, most reads result in cache
            hits.
            % \tristan{Dzung, pls check that.}.
            % Yes, disk reads appear from 23 apps indeed.

@@

    \section{Conclusion}
    \label{discussion}
        Page cache undoubtedly has important impacts on performance, in
        particular for data-intensive applications. Numerical simulation is
        an effective approach to predict and compare the performance of
        applications, schedulers and infrastructures. However, page cache
        is missing in many simulation tools for distributed systems. In
        this study, we proposed a simulation model of I/Os with page cache,
        we implemented it in the \simgrid-based \wrench library, and we
        evaluated its benefits in different conditions. Results show that
        our model very substantially improves simulation accuracy,
        decreasing absolute relative errors between simulated and real
        performance by a factor close to 10 (see results of the
        single-threaded experiment). 
        %Furthermore, this level of accuracy
        %can be achieved without a fine-grained level of details
        %\tristan{not sure what you mean}. 
        % Dzung: I meant the results are accurate without a very detailed model. But the overhead results dot not seem to show that.
        Our page cache model is available
        in \wrench from commit \href{https://github.com/wrench-project/wrench/tree/d4e188d2f0cf69a19ae0a6b5e41530eb77d12cd8}{d4e188d2f0}, ready to be used 
        for the simulation of Big Data applications.

        Simulation results could be made even more accurate by a deeper
        investigation of page cache flushing and eviction order. 
        The availability of asymmetrical bandwidths in the
        forthcoming \simgrid release will also directly improve our
        simulation accuracy. Relevant future work includes simulating
        application memory consumption, random I/O, file readahead and
        persistent disk storage.
